---
title: "Graphics with R"
subtitle: "Smart Analytics for Big Data"
author: "Iragaël Joly"
date: '2020-12-11'
slug: []
categories: ["R"]
tags: ["R Markdown", "plot", "regression"]
authorbox: false
---

# Grammar of `ggplot2()`

## Start example



```{r include=FALSE}
library(tidyverse)
```


```{r}
df <- data.frame(x=1:15, y=1:15)
ggplot(df, aes(x,y)) +
  geom_point() +
  labs(
    title= "Plot example",
    subtitle = "Subtitled",
    x="x-values", y = "y-values"
    ) +
  theme(panel.background = element_rect(fill = "white", color= "grey50"))
```


## Grammar of graphic: `ggplot`

The `ggplot2` package is loaded to access the `ggplot()` function.

The graph is created by stacking together several short phrases, each describing an aspect of the graphic.

With `ggplot`, the data need need not to be reshaped for each type of graphical representation (boxplot, histogram, scatter,...). 

Components of a `ggplot`graph:

### Geometric object functions

Geometric objects that describe the type of graph being created. 

Their names start with `geom_`: `geom_line`, `geom_boxplot`, `geom_point`, etc.

### Aesthetics

Aesthetics mappings communicates to `ggplot` which variables in the source data get mapped to which visual elements in the graphics.

### Stats

Stats are statistical transformations that are done before displaying the data.

Not all graphs will have stats. Examples are `stat_ecdf`, `stat_identity` (which tells to pass the data without doing any stat at all)

### Facet function

Facets are subplots where each plot represents a subgroup of the data.
Functions are `facet_wrap` `facet_grid`.

### Themes

Themes are the visual elements of the graph that are not linked to the data: titles, margins, font, ...

### Layer

A layer is a combination of data, aesthitics, a geometric object, a stat, and other options to produce a visal layer in the graphic.

## *Long* vs *Wide* data with `ggplot`

*Wide* means every variable to be ploted is its own column in the underlying dataframe. This is the approach used in Excel.

R works most easily with *Long* data, where additional data are added as rows in the dataframe. 

The side effect of adding more measurements as rows is that any properly constructed `ggplot`graph will automatically update to reflect the new data, without changing the code.



### Additional references

See *R for Data Science* (chap. 28)  (https://r4ds.had.co.nz/graphics-for-communication.html)[https://r4ds.had.co.nz/graphics-for-communication.html] of H. Wickham and G. Grolemund for additional notes on the graph customisation (labels, position, colours, etc.)




## Examples of graphics

### Creating a Scatter Plot

You want to create a scatter plot of the  paired observations: $(x_1, y_1), (x_2, y_2), \dots, (x_n, y_n)$

```{r eval=FALSE}
ggplot(df, aes(x, y)) +
  geom_point()
```

`ggplot`requires:

- what are the dataframe `df`
- and which is the aesthietic mapping (`aes`) to use. `aes` defines which fiels of `df` goes into which axis of the plot.
- the geometric object: `geom_point` for a point graph.

We can easily add title and axis-label

```{r}
p <- ggplot(mtcars, aes(hp, mpg)) +
  geom_point() +
  labs(title = "Cars: Horsepower vs. Fuel Economy",
       x = "HP",
       y = "Economy (miles per gallon)")
p
```

Some theme options can be used :


```{r}
g1 <- ggplot(df) +
  geom_point(aes(x, y)) +
    labs(title = "Cars: Horsepower vs. Fuel Economy",
       x = "HP",
       y = "Economy (miles per gallon)") +
    theme(panel.background = element_rect(fill = "white", colour = "grey50"))
g1
```


```{r}
g2 <- g1 + theme(panel.grid.major =
                   element_line(color = "black", linetype = 3)) + 
  # linetype = 3 is dash
  theme(panel.grid.minor =
          element_line(color = "darkgrey", linetype = 4)) 
  # linetype = 4 is dot dash
g2
```

### Applying a Theme

Use of a preset collection of colors, styles, and formatting

```{r eval=F}
theme_bw()
theme_dark() 
theme_classic()
theme_gray()
theme_linedraw()
theme_light()
theme_minimal()
theme_test()
theme_void()
```

We can apply each of them. For example:

```{r}
p + theme_bw()
```

Packages, like  `ggtheme`, that include additional themes with some popular theme like *Stata* or *The Economist*


### Scatter Plot of Multiple Groups


The dataframe is multiple observations per record: x, y, and a factor f

```{r}
ggplot(data = iris,
       aes(
         x = Petal.Length,
         y = Petal.Width,
         shape = Species,
         color = Species
       )) +
  geom_point()
```

*About the legend*

It is automatic with a factor in `aes`.

It can be hiden with ` +   theme(legend.position = "none")`

or added with `+ guides(shape=guide_legend(title="My Legend Title")) `

Position can be changed with `+ theme(legend.position = "bottom")` or `+ theme(legend.position = c(.8, .2))` : center of the legend at 80% to the right and 20% up from the bottom


### A scatter plot for each factor values

```{r}
data(Cars93, package = "MASS")
ggplot(Cars93, aes(MPG.city, Horsepower)) +
  geom_point() +
  facet_wrap( ~ Origin)
```


### Regression Line of a Scatter Plot

Adding the regression line with `geom_smooth`:

```{r}
ggplot(df, aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm",
              formula = y ~ x,
              se = FALSE)
```

`se = FALSE` parameter tells ggplot not to plot the standard error bands around our regression 

#### Example

```{r}
library(faraway)
data(strongx)
g <- ggplot(strongx, aes(energy, crossx)) +
  geom_point()

g + geom_smooth(method = "lm", 
                formula = y ~ x)
```

and

```{r}
g + geom_smooth(method = "lm",
                formula = y ~ x,
                se = FALSE)
```

Specific line can be added using `intercept` and `slope`. For example, extractiong the values from a regression

```{r}
m <- lm(crossx ~ energy, data = strongx)

ggplot(strongx, aes(energy, crossx)) +
  geom_point() +
  geom_abline(
    intercept = m$coefficients[1],
    slope = m$coefficients[2]
  )
```

### Plotting All Variables Against All Other Variables

Dataframe contains multiple numeric variables. The scatter plots for all pairs of variables uses the `GGally` package.

```{r}
library(GGally)
ggpairs(iris)
```

or for simpler and faster graph : `plot(iris)`


### Creating a Bar Chart

#### From the count table (or equivalent)

```{r}
ford_cars <- Cars93 %>%
  filter(Manufacturer == "Ford") 

ggplot(ford_cars, aes(Model, Horsepower)) +
  geom_bar(stat = "identity")
```


`stat = "identity"`, which assumes that the heights of your bars are conveniently stored as a value in one field with only one record per column

#### From the 'raw' variable 

From the vector of numeric data and associated factor, you can produce a bar chart of the group means or the group totals

```{r}
ggplot(airquality, aes(month.abb[Month], Temp)) +
  geom_bar(stat = "summary", fun.y = "mean") +
  labs(title = "Mean Temp by Month",
       x = "",
       y = "Temp (deg. F)")
```

#### Sorting the factor in graph

```{r}
library(forcats)

aq_data <- airquality %>%
  arrange(Month) %>%
  mutate(month_abb = fct_inorder(month.abb[Month]))

ggplot(aq_data, aes(month_abb, Temp)) +
  geom_bar(stat = "summary", fun.y = "mean") +
  labs(title = "Mean Temp by Month",
       x = "",
       y = "Temp (deg. F)")
```

Sorted x-axis with respect to the value in y-axis

```{r}
ggplot(aq_data, aes(reorder(month_abb, -Temp, mean), Temp)) +
  geom_bar(stat = "summary",
           fun.y = "mean",
           fill = "tomato") +
  stat_summary(fun.data = mean_se, geom = "errorbar") +
  labs(title = "Mean Temp by Month",
       x = "",
       y = "Temp (deg. F)")
```


Modifiying the color and adding standard error bars:

```{r}
ggplot(aq_data, aes(month_abb, Temp)) +
  geom_bar(stat = "summary",
           fun.y = "mean",
           fill = "cornflowerblue") +
  stat_summary(fun.data = mean_se, geom = "errorbar") +
  labs(title = "Mean Temp by Month",
       x = "",
       y = "Temp (deg. F)")
```

Using a palette color:


```{r}
aq_data <- airquality %>%
  arrange(Month) %>%
  mutate(month_abb = fct_inorder(month.abb[Month]))

ggplot(data = aq_data, aes(month_abb, Temp, fill = month_abb)) +
  geom_bar(stat = "summary", fun.y = "mean") +
  labs(title = "Mean Temp by Month",
       x = "",
       y = "Temp (deg. F)") +
  scale_fill_brewer(palette = "Paired")
```


With colors associated to factor : `fill=`


```{r}
ggplot(airquality, aes(month.abb[Month], Temp, fill = ..y..)) +
  geom_bar(stat = "summary", fun.y = "mean") +
  labs(title = "Mean Temp by Month",
       x = "",
       y = "Temp (deg. F)",
       fill = "Temp")
```


### Plotting a Line from x and y Points

Add the `+ geom_line()` to the scater plot

```{r}
ggplot(economics, aes(date , unemploy)) +
  geom_point() +
  geom_line()
```


#### Changing the Type, Width, or Color of a Line

`linetype=""` parameter for controlling the appearance of lines:

   | `linetype=` Character | `linetype=` Numeric | 
   |-----------------------|---------------------|
   | "solid"   | =1 (default) |
   | "dashed"   | =2 |
   | "dotted"   | =3 |
   | "dotdash"   | =4 |
   | "longdash"   | =5 |
   | "twodash"   | =6 |
   | "blank"   | =0 (inhibits drawing) |


The `linetype`, `col`, and `size` params to `geom_line`:

```{r}
ggplot(df, aes(x, y)) +
  geom_line(linetype = 2,
            size = 2,
            col = "red")
```

#### From wide to long format in Line grah

A classic case is when you want to draw multiple lines with multiple y-variables and a unique x-variable

```{r}
x <- 1:10
y1 <- x**1.5
y2 <- x**2
y3 <- x**2.5
df <- data.frame(x, y1, y2, y3)
```

Pass wide data to  long by using the gather function from the core tidyverse package

```{r}
df_long <- gather(df, bucket, y, -x)
```


```{r}
ggplot(df_long, aes(x, y, col = bucket)) +
  geom_line()
```

#### Plotting Multiple Datasets

creating an empty plot and then adding two different geoms to the plot:

```{r eval=F}
ggplot() +
  geom_line(data = df1, aes(x1, y1)) +
  geom_line(data = df2, aes(x2, y2))
```



```{r}
# example data
n <- 10
n2 <- 21
x1 <- 1:n
y1 <- rnorm(n, 0, .5)
df1 <- data.frame(x1, y1)

x2 <- (.5 * n2):((1.5 * n2) - 1)
y2 <- rnorm(n2, 1, .5)
df2 <- data.frame(x2, y2)
```

```{r}
ggplot() +
  geom_line(data = df1, aes(x1, y1), color = "darkblue") +
  geom_line(data = df2, aes(x2, y2), linetype = "dashed")+
  geom_point(data=df1, aes(x1, y1)) +
  geom_point(data=df2, aes(x2, y2))


```

Adding vertical and horizontal lines


```{r}
ggplot(df1) +
  aes(x = x1, y = y1) +
  geom_point() +
  geom_vline(
    xintercept = 10,
    color = "red",
    linetype = "dashed",
    size = 1.5
     ) +
  geom_hline(yintercept = 0, color = "blue")
```

#### Exercise

```{r include=T}
samp <- rnorm(1000)
samp_df <- data.frame(samp, x = 1:length(samp))

mean_line <- mean(samp_df$samp)
sd_lines <- mean_line + c(-2, -1, +1, +2) * sd(samp_df$samp)

ggplot(samp_df) +
  aes(x = x, y = samp) +
  geom_point() +
  geom_hline(yintercept = mean_line, color = "darkblue") +
  geom_hline(yintercept = sd_lines, linetype = "dotted")
```

### Creating a Boxplot

```{r include=FALSE}
samp <- rnorm(1000)
samp_df <- data.frame(samp, x = 1:length(samp))
```

```{r}
ggplot(samp_df) +
  aes(y = samp) +
  geom_boxplot() +
  coord_flip()
```


```{r}
data(UScereal, package = "MASS")

ggplot(UScereal) +
  aes(x = as.factor(shelf), y = sugars) +
  geom_boxplot() +
  labs(
    title = "Sugar Content by Shelf",
    x = "Shelf",
    y = "Sugar (grams per portion)"
  )
```

#### Creating a Histogram

The `geom_histogram` function must decide how many cells (bins) to create for binning the data.

Automatic bins nunber

```{r}
data(Cars93, package = "MASS")

ggplot(Cars93) +
  geom_histogram(aes(x = MPG.city))
```

Customized bins number

```{r}
ggplot(Cars93) +
  geom_histogram(aes(x = MPG.city), bins = 13)
```

#### Adding a Density Estimate to a Histogram



```{r}
ggplot(Cars93) +
  aes(x = MPG.city) +
  geom_histogram(aes(y = ..density..), bins = 21) +
  geom_density()
```

#### Creating a Normal Quantile–Quantile Plot



```{r}
df <- data.frame(x = rnorm(100))

ggplot(df, aes(sample = x)) +
  stat_qq() +
  stat_qq_line()
```

Example:

```{r echo=F}
p1 <- ggplot(Cars93, aes(sample = Price)) +
  stat_qq() +
  stat_qq_line()
p1
```

Is the $log(price)$ 'more normal'?



### Creating Other Quantile–Quantile Plots

The recipe steps:

  1.  Use the ppoints function to generate a sequence of points between 0 and 1.
  2.  Transform those points into quantiles, using the quantile function for the assumed distribution.
  3.  Sort your sample data.
  4.  Plot the sorted data against the computed quantiles.
  5.  Use abline to plot the diagonal line.

#### Example
Testing the Student distribution

Create the data to be tested

```{r}
df_t <- data.frame(y = rt(100, 5))
```

Estimate the parameters of the distribution we’re wanting to plot. Since this is a Student’s t distribution, we only need to estimate one parameter, the degrees of freedom.
We use `MASS::fitdistr` function

```{r}
est_df <- as.list(MASS::fitdistr(df_t$y, "t")$estimate)[["df"]]
est_df
```

There is a match !

```{r}
p2 <- ggplot(df_t) +
  aes(sample = y) +
  geom_qq(distribution = qt, dparams = est_df) +
  stat_qq_line(distribution = qt, dparams = est_df)
p2
```

### Graphing a Function

`stat_function` will graph a function across a range

```{r}
p3 <- ggplot(data.frame(x = c(-3, 3))) +
  aes(x) +
  stat_function(fun = sin)
p3
```

Another example

```{r}
p4 <- ggplot(data.frame(x = c(-3.5, 3.5))) +
  aes(x) +
  stat_function(fun = dnorm) +
  ggtitle("Standard Normal Density")
p4
```

or again 

```{r}
f <- function(x) exp(-abs(x)) * sin(2 * pi * x)

ggplot(data.frame(x = c(-3.5, 3.5))) +
  aes(x) +
  stat_function(fun = f) +
  ggtitle("Dampened Sine Wave")
```

### Displaying Several Figures on One Page

With the `patchwork` by Thomas Lin Pedersen.

After installing the package, we can use it to plot multiple ggplot objects using a `+` between the objects, then a call to `plot_layout` to arrange the images into a grid, as shown in Figure 10.59. Our example code here has four ggplot objects:

```{r}
#install.packages("patchwork")
library(patchwork)
p1 + p2 + p3 + p4

```

Grouping with parentheses and using / to put groupings under other elements

```{r}
p3 / (p1 + p2 + p4)
```

```{r}
p1 + p2 + p3 + p4 + plot_layout(ncol = 2, byrow = FALSE)
```

### Writing Your Plot to a File

In the  *RStudio* plot panel, the  `Export` button permits to “Save as Image,” “Save as PDF,” or “Copy to Clipboard.” 

The “Copy to Clipboard” option can be handy if you are manually copying and pasting your graphics into a presentation or word processor.

A reproducible manner is to use `ggsave` to save a displayed image to a file. `ggsave` will make some default assumptions about size and file type for you, allowing you to specify only a filename: `ggsave("filename.jpg")`

Remember that the file will be written to your current working directory (unless you use an absolute filepath), so be certain you know which directory is your working directory.

In a noninteractive script using ggplot, you can pass plot objects directly to `ggsave` so they need not be displayed before saving. 

```{r eval=F}
ggsave("g1.png", plot = g1, units = "in", width = 5, height = 4)
```

Note that `in` for inches and `ggsave` also supports `mm` and `cm`

