---
title: "Exercice - Death Penalty"
subtitle: "Inspired from Paul Allison : logistic regression using SAS, 2nd Ed"
author: "I. Joly"
date:  "`r Sys.Date()`"
output: html_document
fontsize: 11pt
link-citations: yes
biblio-style: apalike
bibliography: ["02_biblio.bib"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this exercise special attention is given to misinterpretations. For that the theme is deliberately heavy and sensitive, to force to take precautions before the attive conclusions.

The data is described in @allison2010survival

> Data set consists of 147 death penalty cases in the state of New Jersey. 
The defendant was convicted of first-degree murder with a recommendation by the prosecutor that a death sentence be imposed. 

> Then a penalty trial was conducted to determine whether the defendant would receive a sentence of death or life imprisonment. 

>Our dependent variable DEATH is coded 1 for a death sentence, and 0 for a life sentence. 

>The aim is to determine how this outcome was influenced by various characteristics of the defendant and the crime.

## Chargement des données et packages

```{r message=FALSE, warning=FALSE}
library(psych)  # for summary stat
library(gmodels)  # for xtable
library(DescTools)  # for VIF multicolineartity
library(arm)  # for bayesian estimation
library(knitr)   # for outpout in kable
library(arm)  # for bayesian estimation
library(ResourceSelection)  # FOR hosmer&Lemeshow test

# dataframe has to be present in the rmd folder
DF <- read.csv(file = "02_Penalty.txt", sep = " ", dec = '.', header = T)
```

Variables are:

- `DEATH` : 1 for death sentence / 0 for a life sentence
- `BLACKD`: 1 if the defendant was black, otherwise 0
- `WHITVIC`: 1 if the victim was white, otherwise 0
- `SERIOUS`: a rating of the seriousness of the crime, as evaluated by a panel of four to six judges. (average rankings between 1(least serious)-15(most serious))[^1]

[^1]: `SERIOUS` was developed in an auxiliary study in which panels of trial judges were given written descriptions of each of the crimes. These descriptions did not mention the race of the defendant or the victim

- `CULP`: 5 denotes high culpability and 1 denotes low culpability, based on aggravating and mitigating circumstances)
- `SERIOUS2`: a 5 points rating scale of the seriousness of the crime

### You will discuss the following topics

***

### Descriptive analysis


***

### ChiSquarred test

We can test dependence: the idea that independence between race and sentence is not respected.

***

###  Linear Probability Model

***

### Logistic regression `death ~ blackd + whitvic` 

***

### Logistic regression `death ~ blackd + whitvic+ serious` 

***

### Marginal effects

***

###  Logistic regression `death ~ culp`

***

### Logistic regression `death ~ blackd + whitvic + culp` with `culp:reflevel:5`

***

### Logistic regression `blackd * whitvic + culp`

***

### Hosmer-Lemeshow Test

***

### Multicolinearity Diag


