---
title: "Application of Survival Analysys - Exercise 2"
author: 'IJ'
date: '2020-12-11'
slug: []
categories: []
tags: []
link-citations: yes
biblio-style: apalike
bibliography: [./data/biblio_hdr_v2.bib, ./data/Liste_Publi3.bib, ./data/biblio_SABD.bib]
nocite: |
  @Bivand2013, @lovelace2019
---

# Exercise inspired from @Allison2010

The exercise analyses the job duration `jobdur.txt`. The data consist of 100 job durations, measured from the year to entry into hte job until the year that the employee quit. 

Duration after the fifth year are censored. If the employee is fired, the job duration is censored at the end of its last full year.

Survival times have values of `1, 2, 3, 4 or 5`


## Import data & Summary statistics


```{r}
```{r}
#### Packages
library(dplyr)
library(survival)
library(survminer)
library(lubridate)
library(knitr)
# Data package
library(Ecdat)
##################
jobdur <- read.table("../Lessons/data/jobdur.txt", header = TRUE)
head(jobdur)

summary(jobdur)
addmargins( table(jobdur$dur) )

```

## Data Handling


```{r}
jobdur$censored <- jobdur$event != 1
addmargins(table(jobdur$censored))
addmargins(table(jobdur$event))

Surv_time <- with(jobdur , Surv(dur, censored))
```



## Survival curves & Life Table



1. Simple survival curve that doesn’t consider any different groupings

```{r}
fit <- survfit(Surv_time~1, data = jobdur)
fit
summary(fit)
plot(fit)
```

Or using the package `survminer` and `ggsurvplot()` function:

```{r}
library(survminer)
ggsurvplot(fit)
```


2. Curves depending on education level.

Let use the factor depending on education level $>12$ 

```{r}
ed_level <- factor(jobdur$ed>12, labels=c('Low Educ', 'High Educ') )
```


```{r}
sfit <- survfit(Surv_time~ed_level, data = jobdur)
sfit
summary(sfit)
plot(sfit)
ggsurvplot(sfit, conf.int=TRUE, pval=TRUE, risk.table=TRUE, 
           legend.labs=c("Low Educ", "High Educ"), legend.title="Sex",  
           palette=c("dodgerblue2", "orchid2"), 
           title="KM Curve for Jub Duration Survival", 
           risk.table.height=.15)
```

